---
title: "Metro Vancouver 2016 Census on Housing"
author: "Bo Wen"
date: "2019-10-11"
output: html_document
# output: html_notebook
---

```{r, include=FALSE}
library(tidyverse)
library(sf)
library(rjson)
library(cancensus)
library(leaflet)
library(leafsync)
```


```{r, include=FALSE}
# Read Data ----

# > set up cancensus api key
json_data <- fromJSON(file='secret.json')
options(cancensus.api_key = json_data[["CensusMapper_API_key"]])

# > set up cancensus cache_path 
options(cancensus.cache_path = file.path(getwd(), '_cache_cancensus'))

## Census data
# define census year and spatial level of aggregation
dataset='CA16'
level="DA"
# INFO - level by dissemination area 
# one or more adjacent dissemination blocks with an average population of 400 to 700 persons 
# based on data from the previous Census of Population Program. 
# It is the smallest standard geographic area for which all census data are disseminated.

# > list of all regions
all_regions <- list_census_regions(dataset)
#select GVRD for analysis
regions <- all_regions %>%
  filter(level==level) %>% 
  filter(CD_UID %in% c(5915)) %>% # Metro Van
  as_census_region_list

# > list of all vectors
all_vectors <- list_census_vectors(dataset)

# # > find variables
# all_vectors %>%
#   filter(grepl("income",label)) %>%
#   # filter(vector == 'v_CA16_406') %>%
#   View

# > variable list
# v_CA16_4895	Total	Median value of dwellings ($)
# v_CA16_4893	Total	Median monthly shelter costs for owned dwellings ($)
# v_CA16_4900	Total	Median monthly shelter costs for rented dwellings ($)
#
# v_CA16_2397	Total	Median total income of households in 2015 ($)
# v_CA16_2207	Total	Median total income in 2015 among recipients ($)
# v_CA16_2213	Total	Median after-tax income in 2015 among recipients ($)
# v_CA16_2219	Total	Median market income in 2015 among recipients ($)
# 
# v_CA16_401	Total	Population, 2016
# v_CA16_407	Total	Land area in square kilometres
# v_CA16_406	Total	Population density per square kilometre
# v_CA16_379	Total	Average age
# v_CA16_425	Total	Average household size
# v_CA16_2449	Total	Average family size of economic families
# v_CA16_5603	Total	Employed
# v_CA16_5651	Total	Self-employed

# set vectors
vectors <- all_vectors %>%
  filter(vector %in% c(
    "v_CA16_4895",
    "v_CA16_4893",
    "v_CA16_4900",
    "v_CA16_2397",
    "v_CA16_401",
    "v_CA16_407",
    "v_CA16_406",
    "v_CA16_379",
    "v_CA16_425",
    "v_CA16_2449",
    "v_CA16_5603",
    "v_CA16_5651"
  )) %>%
  pull("vector")

# > get the data
census_data_all <- get_census(dataset = dataset,
                          level=level,
                          vectors=vectors, 
                          regions=regions, 
                          geo_format = "sf",
                          use_cache=TRUE)

# data preprocessing
census_data <- census_data_all %>%
  # keep only DA
  filter(Type == 'DA') %>%
  # select rename variable columns
  select(
    GeoUID,
    RegionName = "Region Name",
    Type,
    Dwellings,
    Households,
    Population,
    shelter_val_med = "v_CA16_4895: Median value of dwellings ($)",
    own_cost_med = "v_CA16_4893: Median monthly shelter costs for owned dwellings ($)",
    rent_cost_med = "v_CA16_4900: Median monthly shelter costs for rented dwellings ($)",
    pop_tot = "v_CA16_401: Population, 2016",
    pop_density = "v_CA16_406: Population density per square kilometre",
    land_area = "v_CA16_407: Land area in square kilometres",
    hhsize_avg = "v_CA16_425: Average household size",
    age_avg = "v_CA16_379: Average age",
    # income_tot_med = "v_CA16_2207: Median total income in 2015 among recipients ($)",
    # income_at_med = "v_CA16_2213: Median after-tax income in 2015 among recipients ($)",
    # income_mk_med = "v_CA16_2219: Median market income in 2015 among recipients ($)",
    income_hh_med = "v_CA16_2397: Median total income of households in 2015 ($)",
    famsize_avg = "v_CA16_2449: Average family size of economic families",
    emp_tot = "v_CA16_5603: Employed",
    self_emp_tot = "v_CA16_5651: Self-employed",
    land_area_shp =  "Area (sq km)"
  ) %>%
  mutate(
    shelter_cost_med = max(own_cost_med, rent_cost_med),
    emp_density = (emp_tot + self_emp_tot) / land_area,
    pop_emp_density = (pop_tot + emp_tot + self_emp_tot) / land_area
  ) %>%
  filter(pop_tot > 250) %>%
  filter(pop_density < 100e3) %>%
  filter(famsize_avg != 0) %>%
  mutate(
    own_cost_med = if_else(own_cost_med == 0, NA_real_, own_cost_med),
    rent_cost_med = if_else(rent_cost_med == 0, NA_real_, rent_cost_med),
    shelter_val_med = if_else(shelter_val_med == 0, NA_real_, shelter_val_med)
  ) %>%
  mutate(housing_income_ratio = shelter_val_med / income_hh_med) 

```


# Data Exploratory Analysis

There are 3,400 Census dissemination areas in Metro Vancouver, ranging from 0.01 to 145 km^2 in land areas. This census area accounts for over 955,000 households and 245,000 people.

```{r, echo=FALSE,warning=FALSE}

# Data EDA ----
# > map area ----
## Map of Metro Vancouver Census Dissemination Areas
# # map of shapes
# census_data %>%
#   leaflet() %>%
#   addProviderTiles(providers$CartoDB.Positron) %>%
#   addPolylines(color = 'green', weight = 2)

```


## Map of Key Variables on Housing by Dissemination Areas

We extracted some key relevant variables on housing, by dissemination area, such as:
* Medium Shelter Value, 
* Medium Total Household Income, 
* Average Household Size, and 
* Population Density

*Mouse over the map for values of the variables for each area*

You can examine these variables on a series of interactive map below:

```{r, echo=FALSE,warning=FALSE}

# build a template for maps
## common label
labels <- sprintf(
  "<strong>%s %s</strong><br/>Med. Shelter Value: $%gK <br/>Med. Income: $%gK <br/>House Price-Income Ratio: %g <br/>Avg. HH Size: %g <br/>Avg. Pop. Density: %g",
  census_data$RegionName,
  census_data$GeoUID,
  round(census_data$shelter_val_med,-3) / 1e3,
  round(census_data$income_hh_med,-3) / 1e3,
  census_data$housing_income_ratio,
  census_data$hhsize_avg,
  census_data$pop_density
) %>% lapply(htmltools::HTML)

# common map template
tidy_lf_map <-  function(df,
                         var,
                         var_name,
                         pal,
                         pal_var,
                         label,
                         labFormat = labelFormat(),
                         ...) {
  lf_map <- df %>%
    leaflet() %>%
    addProviderTiles(providers$CartoDB.Positron) %>%
    addPolygons(
      fillColor = ~ pal_var,
      opacity = 1,
      stroke = TRUE,
      weight = 0.5,
      color = "white",
      dashArray = "",
      fillOpacity = 0.7,
      highlight = highlightOptions(
        weight = 2,
        color = "#666",
        dashArray = "",
        fillOpacity = 0.7,
        bringToFront = TRUE
      ),
      label = label,
      labelOptions = labelOptions(
        style = list("font-weight" = "normal", padding = "3px 8px"),
        textsize = "15px",
        direction = "auto"
      )
    ) %>%
    addLegend(
      "bottomright",
      pal = pal,
      values = ~ var,
      title = var_name,
      labFormat = labFormat,
      opacity = 1
    )
  return(lf_map)
}

# map shelter value
bins <- c(0, 100e3, 500e3, 1e6, 1.5e6, 2e6, 2.5e6, 3e6, 5e6)
pal1 <- colorBin("Reds", domain = census_data$shelter_val_med, bins = bins)
shelter_map <-  census_data %>% tidy_lf_map(
  var = .$shelter_val_med,
  var_name = "Med. Housing Value",
  pal = pal1,
  pal_var = pal1(.$shelter_val_med),
  label = labels,
  labFormat = labelFormat(prefix = "$"),
)

# > map total HH income value
bins <- c(0, 25e3, 50e3, 75e3, 100e3, 125e3, 150e3, 200e3, 300e3)
pal1 <- colorBin("YlGn", domain = census_data$income_hh_med, bins = bins)
income_map <-  census_data %>% tidy_lf_map(
  var = .$income_hh_med,
  var_name = "Med. Household Income",
  pal = pal1,
  pal_var = pal1(.$income_hh_med),
  label = labels,
  labFormat = labelFormat(prefix = "$"),
)

# map house price to income ratio
bins <- c(0, 2, 4, 6, 8, 10, 20, 50, 100)
pal1 <- colorBin("YlOrRd", domain = census_data$housing_income_ratio, bins = bins)
housing_income_map <-  census_data %>% tidy_lf_map(
  var = .$housing_income_ratio,
  var_name = "Housing-income Ratio",
  pal = pal1,
  pal_var = pal1(.$housing_income_ratio),
  label = labels
  # labFormat = labelFormat(prefix = "$"),
)

# map HH size
bins <- c(1, 1.5, 2, 2.5, 3, 3.5, 4, 4.5, 5)
pal1 <- colorBin("YlGnBu", domain = census_data$hhsize_avg, bins = bins)
avgHHSize <-  census_data %>% tidy_lf_map(
  var = .$hhsize_avg,
  var_name = "Household Size",
  pal = pal1,
  pal_var = pal1(.$hhsize_avg),
  label = labels
  # labFormat = labelFormat(prefix = "$"),
)

# map Pop Density
bins <- c(10, 1e3, 2.5e3, 5e3, 10e3, 25e3, 50e3, 80e3)
pal1 <- colorBin("BuPu", domain = census_data$pop_density, bins = bins)
popDensity <-  census_data %>% tidy_lf_map(
  var = .$pop_density,
  var_name = "Population Density",
  pal = pal1,
  pal_var = pal1(.$pop_density),
  label = labels
  # labFormat = labelFormat(prefix = "$"),
)

latticeView(
  # shelter_map,
  housing_income_map,
  # avgHHSize,
  popDensity,
  sync = "all",
  ncol = 1
)

```

&nbsp;

While the interactive maps provide us with a general impression about how housing values, income, household size and population density can vary across the Metro Vancouver region, it is difficult draw correlations between housing value and different variables. Let's look at this closer in the next section of the EDA.


## Relationships between key variables and Housing Value

**The relationship between population density and housing value is nonlinear. Higher population density areas may have smaller units of housing that are less expensive.**

```{r, echo=FALSE,warning=FALSE}

# > graphing ----

# housing value vs pop density
ggplot(census_data, aes(pop_density, shelter_val_med)) +
  geom_point() +
  stat_smooth(method = lm)

# housing value vs log_e pop density
ggplot(census_data, aes(1/pop_density, shelter_val_med)) +
  geom_point() +
  stat_smooth(method = lm)

```

**The relationship between housing value and total population in the area seems to be inversely proportional.**

```{r, echo=FALSE,warning=FALSE}

# > graphing ----

# housing value vs total pop
ggplot(census_data, aes(pop_tot, shelter_val_med)) +
  geom_point() +
  stat_smooth(method = lm)

# housing value vs log_e pop density
ggplot(census_data, aes(1/pop_tot, shelter_val_med)) +
  geom_point() +
  stat_smooth(method = lm)

```

**Housing value generally increase as the medium household income in the dissemination area increases.**

```{r, echo=FALSE,warning=FALSE}

# housing value vs hh income
ggplot(census_data, aes(income_hh_med, shelter_val_med)) +
  geom_point() +
  stat_smooth(method = lm)

```


**A larger household also correlates with higher housing value, this can be potentially due to the need for more space and a bigger house.**


```{r, echo=FALSE,warning=FALSE}

# housing value vs hh size
ggplot(census_data, aes(hhsize_avg, shelter_val_med)) +
  geom_point() +
  stat_smooth(method = lm)

# # housing value vs total employment
# ggplot(census_data, aes(emp_tot, shelter_val_med)) +
#   geom_point() +
#   stat_smooth(method = lm)
# ggplot(census_data, aes(1/emp_tot, shelter_val_med)) +
#   geom_point() +
#   stat_smooth(method = lm)
# 
# # housing value vs other variables
# ggplot(census_data, aes(housing_income_ratio, shelter_val_med)) +
#   geom_point() +
#   stat_smooth(method = lm)
# ggplot(census_data, aes(own_cost_med, shelter_val_med)) +
#   geom_point() +
#   stat_smooth(method = lm)
# ggplot(census_data, aes(rent_cost_med, shelter_val_med)) +
#   geom_point() +
#   stat_smooth(method = lm)
# ggplot(census_data, aes(rent_cost_med, shelter_val_med)) +
#   geom_point() +
#   stat_smooth(method = lm)

```


# A Simple Regression Model

```{r, include=FALSE}
# Simple model of housing value by DA ----
census_data_rev <- census_data %>%
  mutate(inv_pop_density = 1 / pop_density,
         inv_emp_tot = 1 / emp_tot,
         inv_pop_tot = 1/ pop_tot)
```

## Model 1

```{r, echo=TRUE,warning=FALSE}
housing_value_model <-
  lm(shelter_val_med ~ income_hh_med + hhsize_avg + inv_pop_density + inv_emp_tot,
     data = census_data_rev)
summary(housing_value_model, corr = T)

```

## Model 2

```{r, echo=TRUE,warning=FALSE}
housing_value_model <-
  lm(shelter_val_med ~ income_hh_med + hhsize_avg + inv_pop_tot + inv_emp_tot,
     data = census_data_rev)
summary(housing_value_model, corr = T)

```


**What do you think of these two models? What other variables you think we should be looking at?**
